---
title: "Pie Charts"

author: "TERE"

date: "2024-06-21"

categories: ["R Graphics", "R Programming"]

tags: ["R", "graphics", "data visualization", "pie charts"]

description: "Learn how to create and customize pie charts in R. This lecture covers essential techniques for visualizing categorical data using pie charts in R."
---



## Introduction



Pie charts are a common way to visualize the proportions of different categories in a dataset. They represent data as slices of a circle, with each slice corresponding to a category's proportion. In this lecture, we will learn how to create and customize pie charts in R.



## Key Concepts



### 1. What is a Pie Chart?



A pie chart is a circular graph divided into slices to illustrate numerical proportions. Each slice's size is proportional to the value it represents.



### 2. When to Use Pie Charts



Pie charts are best used for:

- Displaying the proportions of categorical data.

- Comparing parts of a whole.



### 3. Customizing Pie Charts



Customizing pie charts involves adding titles, labels, colors, and legends to enhance readability and interpretability.



## Creating and Customizing Pie Charts



### 1. Basic Pie Chart



A basic pie chart displays the proportions of different categories.



```{r echo=TRUE}

# Creating sample data

categories <- c("A", "B", "C", "D")

counts <- c(23, 45, 12, 56)



# Creating a basic pie chart

pie(counts, labels = categories, main = "Basic Pie Chart")

```





### 2. Adding Colors



You can add colors to the pie chart to differentiate between categories.



```{r echo=TRUE}

# Creating a pie chart with colors

pie(counts, labels = categories, main = "Pie Chart with Colors", col = rainbow(length(categories)))

```





### 3. Adding Percentages



Adding percentages to the pie chart slices provides more information about the proportions.



```{r echo=TRUE}

# Calculating percentages

percentages <- round(counts / sum(counts) * 100)

labels <- paste(categories, percentages, "%", sep = " ")



# Creating a pie chart with percentages

pie(counts, labels = labels, main = "Pie Chart with Percentages", col = rainbow(length(categories)))

```






### 4. Adding Legends



Adding a legend helps in identifying the categories more easily.



```{r echo=TRUE}

# Creating a pie chart with legend

pie(counts, labels = categories, main = "Pie Chart with Legend", col = rainbow(length(categories)))

legend("topright", legend = categories, fill = rainbow(length(categories)))

```






### 5. Exploding Slices



Exploding slices (pulling them out from the center) can highlight specific categories.



```{r echo=TRUE}

# Creating a pie chart with exploded slices

pie(counts, labels = categories, main = "Pie Chart with Exploded Slices", col = rainbow(length(categories)), explode = 0.1)

```



```{r}

# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Check if plotrix is installed, if not, install it
if (!requireNamespace("plotrix", quietly = TRUE)) {
  install.packages("plotrix")
}

# Load plotrix library
library(plotrix)

# Creating sample data (make sure these are defined earlier in your document)
categories <- c("A", "B", "C", "D")
counts <- c(23, 45, 12, 56)

# Creating a pie chart with exploded slices using the plotrix package
pie3D(counts, labels = categories, main = "Pie Chart with Exploded Slices", 
      col = rainbow(length(categories)), explode = 0.1)

```



## Example: Comprehensive Pie Chart Analysis



Here's a comprehensive example of creating and customizing pie charts in R.



```{r echo=TRUE}

# Creating sample data

categories <- c("A", "B", "C", "D")

counts <- c(23, 45, 12, 56)



# Basic pie chart

pie(counts, labels = categories, main = "Basic Pie Chart")



# Pie chart with colors

pie(counts, labels = categories, main = "Pie Chart with Colors", col = rainbow(length(categories)))



# Pie chart with percentages

percentages <- round(counts / sum(counts) * 100)

labels <- paste(categories, percentages, "%", sep = " ")

pie(counts, labels = labels, main = "Pie Chart with Percentages", col = rainbow(length(categories)))



# Pie chart with legend

pie(counts, labels = categories, main = "Pie Chart with Legend", col = rainbow(length(categories)))

legend("topright", legend = categories, fill = rainbow(length(categories)))



# Pie chart with exploded slices using the plotrix package

# Install the 'plotrix' package if not already installed

install.packages("plotrix")

library(plotrix)

pie3D(counts, labels = categories, main = "Pie Chart with Exploded Slices", col = rainbow(length(categories)), explode = 0.1)

```



## Summary



In this lecture, we covered how to create and customize pie charts in R. We explored various techniques for adding colors, percentages, legends, and exploding slices. Pie charts are a useful tool for visualizing categorical data and understanding the proportions of different categories.



## Further Reading



For more detailed information, consider exploring the following resources:



- [Pie Charts in R](https://www.datamentor.io/r-programming/pie-chart/)

- [R Documentation on pie](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/pie)

- [R for Data Science](https://r4ds.had.co.nz/data-visualization.html)



## Call to Action



If you found this lecture helpful, make sure to check out the other lectures in the R Graphs series. Happy plotting!